Place of death in NRS COVID-19 deaths

### 2021-02-12 covid_nrs_place_of_death.Rmd
figured out how to keep track of data being updated - there's now a separate script that downloads file and keeps track of latest
now I need to modify the .Rmd to read the downloaded data instead!

---
title: "Place of death in NRS COVID-19 deaths"
author: "Jan Savinc"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: html_document
editor_options: 
  chunk_output_type: console
---

@@

@@

```{r}
# checking out structure

@@

To get council areas in Scotland we need NUTS 2 level map data.q

@@

## Scottish LAU 2011 level 1
# curl_download(url = "https://borders.ukdataservice.ac.uk/ukborders/easy_download/prebuilt/shape/Scotland_laulevel1_2011.zip", destfile = zip_file, quiet = FALSE)  # download file from ONS
# unzip(zipfile = zip_file, exdir = unzipped)
# name <- list.files(unzipped, pattern=".shp")
# shapefile <- st_read(file.path(unzipped, name))

@@

# merged_deaths_la %>%
#   left_join(shapefile,by=c("ref_area"="name")) %>%
#   ggplot(aes(fill = deaths_mean, geometry = geometry)) +
#   geom_sf()
#

@@

@@

```{r, fig.width=12, fig.height=9, warning=FALSE}
# TODO: add start of lockdowns 1-3?
# TODO: what is that dip at week 38/39? not obvious - start of september was tier 4 restrictions in big LAs
figure_deaths_at_home <-
  merged_data_overall %>%
  filter(sex == "all" & age == "all") %>%
  filter(place_of_death %in% c("Home & other non-institution")) %>%
  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid), 
               names_to = "deaths_type", values_to = "deaths") %>% 
  ggplot(aes(x = week_number_run_over)) +
  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max), fill = "#4477aa", alpha = 0.5) +
  geom_line(aes(y = deaths, colour = deaths_type)) +
  geom_line(aes(y = deaths_mean), linetype = "dashed", size = 0.2) +
  geom_vline(xintercept = 54, linetype = "dotted") +
  scale_x_continuous(breaks = seq(0,60,by=10)) +
  facet_grid(~place_of_death) +
  theme_minimal() +
  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
                      labels = c("All deaths", 
                                 "Non Covid-19 deaths"),
                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
                      ) +
  theme(legend.position = "top", plot.caption = element_text(size = 10, colour = "gray60")) +
  labs(x = "Week number",
       y = "Deaths (N)",
       subtitle = date_range_text,
       caption = paste0(c(vertical_line_text, source_nrs_text), collapse="\n"),
       colour = "Cause of death"
       )

@@

>>>>

---
title: "Place of death in NRS COVID-19 deaths"
author: "Jan Savinc"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: console
params:
  data_sources: '`r tibble()`'
---

@@

# Dates represent event registration, not occurrence!

From NRS website:

> All routine vital events information we publish is based on the date of registration, not the date on which the event occurred. Inevitably, of course, there are delays between the occurrence of an event and its registration.

@@

```{r}
download_dir <- "./downloaded_data"
if (!dir.exists(download_dir)) dir.create(download_dir)  # create download dir if it doesn't exist

# checking out structure

@@

TODO: figure out which boundaries in Scotland correspond to council areas!

~~To get council areas in Scotland we need NUTS 2 level map data~~, and the most appropriate level of detail to use is likely the BUC - Ultra Generalised Clipped

https://geoportal.statistics.gov.uk/datasets/nuts-level-2-january-2018-ultra-generalised-clipped-boundaries-in-the-united-kingdom

https://geoportal.statistics.gov.uk/datasets/nuts-level-2-january-2018-ultra-generalised-clipped-boundaries-in-the-united-kingdom?geometry=-12.122%2C55.353%2C4.566%2C57.479

health board boundaries April 2019, BUC
https://opendata.arcgis.com/datasets/def40bdc98a9457aa108eb3a5fb052b1_0.zip?outSR=%7B%22latestWkid%22%3A27700%2C%22wkid%22%3A27700%7D

getting error trying to use shapefile with ggplot:
> Error in CPL_transform(x, crs, aoi, pipeline, reverse) : 
  crs not found: is it missing?

explicitly setting crs on shapefile fails,

version fo sf 0.9-7:
> Linking to GEOS 3.8.0, GDAL 3.0.4, PROJ 6.3.1

@@

## Scottish LAU 2011 level 1
# curl_download(url = "https://borders.ukdataservice.ac.uk/ukborders/easy_download/prebuilt/shape/Scotland_laulevel1_2011.zip", destfile = zip_file, quiet = FALSE)  # download file from ONS
# name <- list.files(unzipped, pattern=".shp")
# shapefile <- st_read(file.path(unzipped, name))

@@

download_and_import_shapefile <- function(url) {
  zip_file <- tempfile()
  unzipped <- tempfile()
  downloaded_zip <- curl_download(url = url, destfile = zip_file, quiet = FALSE) 
  unzip(zipfile = zip_file, exdir = unzipped)
  name <- list.files(unzipped, pattern=".shp$")
  shapefile <- st_read(file.path(unzipped, name))
  return(shapefile)
}

health_boards_shapefile_url <- "https://maps.gov.scot/ATOM/shapefiles/SG_NHS_HealthBoards_2019.zip"
la_shapefile_url <- "https://opendata.arcgis.com/datasets/3a4fa2ce68f642e399b4de07643eeed3_0.zip?outSR=%7B%22latestWkid%22%3A27700%2C%22wkid%22%3A27700%7D"

shapefile_hb <- download_and_import_shapefile(url = health_boards_shapefile_url)
shapefile_la <- download_and_import_shapefile(url = la_shapefile_url)

# shapefile %>%
#   ggplot(aes(geometry = geometry)) +
#   geom_sf()

# shapefile_la %>%
#   inner_join(merged_deaths_la, by = c("lad19nm"="ref_area"))

# shapefile_hb %>%
#   inner_join(merged_deaths_hb, by = c("HBName"="ref_area"))

# TODO: summarise data by LA/HB so it can be shown on a map
# otherwise it loads forever because we're re-plotting the same regional boundary for each row!!

merged_deaths_la %>%
  filter(week_number==53 & place_of_death=="Home & other non-institution") %>%
  left_join(shapefile_la,by=c("ref_area"="lad19nm")) %>%
  ggplot(aes(fill = deaths_all_causes, geometry = geometry)) +
  geom_sf()

# ggplot(data = shapefile_hb, aes(geometry=geometry)) + geom_sf()

# merged_deaths_hb %>%
#   left_join(shapefile_hb,by=c("ref_area"="HBName")) %>%
#   ggplot(aes(fill = deaths_mean, geometry = geometry)) +
#   geom_sf()

@@

death_registrations_text <- "Note: all figures are based on date of registration."

@@

```{r, fig.width=12, fig.height=9, warning=FALSE}
# TODO: add start of lockdowns 1-3?
# TODO: what is that dip at week 38/39? not obvious - start of september was tier 4 restrictions in big LAs
figure_deaths_at_home <-
  merged_data_overall %>%
  filter(sex == "all" & age == "all") %>%
  filter(place_of_death %in% c("Home & other non-institution")) %>%
  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid), 
               names_to = "deaths_type", values_to = "deaths") %>% 
  ggplot(aes(x = week_number_run_over)) +
  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max), fill = "#4477aa", alpha = 0.5) +
  geom_line(aes(y = deaths, colour = deaths_type)) +
  geom_line(aes(y = deaths_mean), linetype = "dashed", size = 0.2) +
  geom_vline(xintercept = 54, linetype = "dotted") +
  scale_x_continuous(breaks = seq(0,60,by=10)) +
  facet_grid(~place_of_death) +
  theme_minimal() +
  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
                      labels = c("All deaths", 
                                 "Non Covid-19 deaths"),
                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
                      ) +
  theme(legend.position = "top", plot.caption = element_text(size = 10, colour = "gray60")) +
  labs(x = "Week number",
       y = "Deaths (N)",
       subtitle = date_range_text,
       caption = paste0(c(vertical_line_text, death_registrations_text, source_nrs_text), collapse="\n"),
       colour = "Cause of death"
       )

@@

## Deaths at home in 2020, all of Scotland

```{r, fig.width=12, fig.height=9, warning=FALSE}
figure_deaths_at_home_2020 <-
  merged_data_overall %>%
  filter(sex == "all" & age == "all") %>%
  filter(place_of_death %in% c("Home & other non-institution")) %>%
  filter(year == 2020) %>%
  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid), 
               names_to = "deaths_type", values_to = "deaths") %>% 
  ggplot(aes(x = week_number_run_over)) +
  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max), fill = "#4477aa", alpha = 0.5) +
  geom_line(aes(y = deaths, colour = deaths_type)) +
  geom_line(aes(y = deaths_mean), linetype = "dashed", size = 0.2) +
  scale_x_continuous(breaks = seq(0,60,by=10)) +
  facet_grid(~place_of_death) +
  theme_minimal() +
  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
                      labels = c("All deaths", 
                                 "Non Covid-19 deaths"),
                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
                      ) +
  theme(legend.position = "top", plot.caption = element_text(size = 10, colour = "gray60")) +
  labs(x = "Week number",
       y = "Deaths (N)",
       # subtitle = date_range_text,
       subtitle = "Data from 2020 shown, starting with w/c 30 Dec 2019.",
       caption = paste0(c(death_registrations_text, source_nrs_text), collapse="\n"),
       colour = "Cause of death"
       )

walk(
  .x = c(".pdf",".svg",".png"),
  .f = ~ggsave(filename = paste0("./outputs/figure_deaths_at_home_2020",.x), plot = figure_deaths_at_home_2020, width = 12, height = 9, dpi = 300)
)

figure_deaths_at_home_2020
```
